---
title: "Data Visualization and Batch Correction"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

The previous step is [basic data wrangling](100_basic_data_wrangling.html).

### Hippocampus Proteome Visualization

Loading in clean data.

```{r load_hippo_proteome}
adbxd_hippo_annot =        readRDS("./data/adbxd_hippo_proteome_annotation.RDS")
adbxd_hippo_proteome =     readRDS("./data/adbxd_hippo_proteome_data.RDS")
adbxd_hippo_metadata =     readRDS("./data/adbxd_hippo_proteome_metadata.RDS")
adbxd_hippo_metadata_jax = readRDS("./data/adbxd_hippo_proteome_jax_metadata.RDS")
adbxd_hippo_metadata_uw =  readRDS("./data/adbxd_hippo_proteome_uw_metadata.RDS")
```

Doing some manipulation of metadata.

```{r hippo_proteome_metadata}
library("tidyverse")
library("lubridate")
adbxd_hippo_metadata = adbxd_hippo_metadata |>
  mutate(Sex = gsub("^([MF])[ae].*$", "\\1", condition),
         Age = gsub("^([MF][ae]\\w+)-([OY]).*-([WF]).*$", "\\2", condition),
         Genotype = gsub("^([MF][ae]\\w+)-([OY]).*-([WF]).*$", "\\3", condition),
         MassSpec_date = gsub("^(.*)-Lumos.*$","\\1",ms_filename)) |>
  mutate(Sex = ifelse(Sex == "Hippocampus reference pool" | 
                      Sex == "Mouse brain reference pool", 
                      NA, Sex),
         Age = ifelse(Age == "Hippocampus reference pool" | 
                      Age == "Mouse brain reference pool", 
                      NA, Age),
         Genotype = ifelse(Genotype == "Hippocampus reference pool" | 
                           Genotype == "Mouse brain reference pool", 
                           NA, Genotype)) |>
  mutate(Sex = ifelse(Sex == "M", "Male", Sex),
         Age = ifelse(Age == "Y", "Months06", Age),
         Genotype = ifelse(Genotype == "F", "5xFAD", Genotype)) |>
  mutate(Sex = ifelse(Sex == "F", "Female", Sex),
         Age = ifelse(Age == "O", "Months14", Age),
         Genotype = ifelse(Genotype == "W", "Ntg", Genotype),
         MassSpec_date = dmy(MassSpec_date))
```

Getting value to filter low-expressed proteins.

```{r get_low_expressed_n_hippo}
sample_size = adbxd_hippo_metadata |> 
  filter(!is.na(Sex)) |> 
  group_by(Sex, Age, Genotype) |> 
  summarize(n = length(age_months))

min_n = floor(sum(sample_size$n) * 0.75)
```

Visualizing low-dimensional projection of hippocampus proteome.

```{r visualize_hippo_proteome}
library("MASS")
library("limma")
hippocampus_cols = grep("^HC2_", colnames(adbxd_hippo_proteome))
expression_filter = row.names(adbxd_hippo_proteome[which(rowSums(adbxd_hippo_proteome[,hippocampus_cols] > 0) >= min_n),])
adbxd_hippo_dist = dist(t(adbxd_hippo_proteome[expression_filter,hippocampus_cols]))
adbxd_hippo_isomds = isoMDS(adbxd_hippo_dist)
adbxd_hippo_pca = prcomp(t(adbxd_hippo_proteome[expression_filter,hippocampus_cols]),
                         scale = TRUE, center = TRUE)
adbxd_hippo_mds = limma::plotMDS(adbxd_hippo_proteome[expression_filter,hippocampus_cols], plot = FALSE)
adbxd_hippo_dimreduct = as.data.frame(adbxd_hippo_isomds)
colnames(adbxd_hippo_dimreduct) = gsub("\\.","_",colnames(adbxd_hippo_dimreduct))
adbxd_hippo_dimreduct$PC1 = adbxd_hippo_pca$x[row.names(adbxd_hippo_dimreduct),1]
adbxd_hippo_dimreduct$PC2 = adbxd_hippo_pca$x[row.names(adbxd_hippo_dimreduct),2]
adbxd_hippo_dimreduct$PC3 = adbxd_hippo_pca$x[row.names(adbxd_hippo_dimreduct),3]
adbxd_hippo_dimreduct$PC4 = adbxd_hippo_pca$x[row.names(adbxd_hippo_dimreduct),4]

adbxd_hippo_dimreduct$MD1 = adbxd_hippo_mds$x
adbxd_hippo_dimreduct$MD2 = adbxd_hippo_mds$y
adbxd_hippo_dimreduct$Sex = adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"Sex"]
adbxd_hippo_dimreduct$Age = adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"Age"]
adbxd_hippo_dimreduct$Genotype = adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"Genotype"]
adbxd_hippo_dimreduct$Strain = adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"strain"]
adbxd_hippo_dimreduct$MassSpec_Date = adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"MassSpec_date"]
adbxd_hippo_dimreduct$Batch = factor(adbxd_hippo_metadata[row.names(adbxd_hippo_dimreduct),"batch"],
                                     levels = as.character(1:10),
                                     ordered = TRUE)
adbxd_hippo_dimreduct = adbxd_hippo_dimreduct |>
  arrange(MassSpec_Date, Batch)
# adbxd_hippo_dimreduct$rbatch = sample(adbxd_hippo_dimreduct$Batch)
# adbxd_hippo_dimreduct$is_pool = is.na(adbxd_hippo_dimreduct$Sex)

# adbxd_hippo_dimreduct = adbxd_hippo_dimreduct[which(!is.na(adbxd_hippo_dimreduct$Sex)),]

library("ggplot2")
ggplot(data = adbxd_hippo_dimreduct, 
       aes(x = points_1, y = points_2, col = Batch)) +
  geom_point() +
  ggtitle("Hippocampus Proteome: MDS Plot Shows Batch Effects") +
  xlab("MD1") + ylab("MD2") +
  scale_color_manual(values = scales::hue_pal()(length(levels(adbxd_hippo_dimreduct$Batch)))) +
  theme_bw() +
  theme(panel.grid = element_line(color = "#FFFFFF"))
  
```

There is a batch effect.

```{r visualize_hippo_proteome_date_effect}
ggplot(data = adbxd_hippo_dimreduct, 
       aes(x = points_1, y = points_2, col = as.character(MassSpec_Date))) +
  geom_point() +
  ggtitle("Hippocampus Proteome: MDS Plot Shows Batch Effects") +
  xlab("MD1") + ylab("MD2") +
  scale_color_manual(name = "Date", 
                     values = scales::hue_pal()(length(unique(adbxd_hippo_dimreduct$MassSpec_Date)))) +
  theme_bw() +
  theme(panel.grid = element_line(color = "#FFFFFF"))
  
```

The batch effect appears somewhat -- but not solely -- related to date. Visualizing the batch effect with boxplots.

```{r full_boxplot}
adbxd_hippo_proteome_for_boxplot = as.data.frame(adbxd_hippo_proteome[expression_filter,hippocampus_cols])
adbxd_hippo_proteome_for_boxplot$protein = row.names(adbxd_hippo_proteome_for_boxplot)
adbxd_hippo_proteome_for_boxplot_long = adbxd_hippo_proteome_for_boxplot |>
  pivot_longer(cols = starts_with("HC2"),
               names_to = "sample",
               values_to = "expression") |>
  mutate(sample = factor(sample, levels = row.names(adbxd_hippo_dimreduct), ordered = TRUE))
adbxd_hippo_proteome_for_boxplot_long$Batch = adbxd_hippo_dimreduct[adbxd_hippo_proteome_for_boxplot_long$sample,"Batch"]
adbxd_hippo_proteome_for_boxplot_long$MassSpec_Date = as.character(adbxd_hippo_dimreduct[adbxd_hippo_proteome_for_boxplot_long$sample,"MassSpec_Date"])

ggplot(data = adbxd_hippo_proteome_for_boxplot_long, aes(x = sample, y = expression, color = Batch)) +
  geom_boxplot() +
  xlab(NULL) +
  theme_bw() +
  scale_color_manual(values = scales::hue_pal()(length(unique(adbxd_hippo_proteome_for_boxplot_long$Batch)))) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = "none")
```

Visualizing the batch effect in the reference hippocampus and brain pools.

```{r visualize_batch_hippo_pool}
library("tidyverse")
mouse_hippo_pool_samples = row.names(adbxd_hippo_metadata[which(adbxd_hippo_metadata$condition == "Hippocampus reference pool"),])
mouse_hippo_pool = adbxd_hippo_proteome[,mouse_hippo_pool_samples]
mouse_hippo_pool = mouse_hippo_pool[which(rowSums(mouse_hippo_pool != 0) == ncol(mouse_hippo_pool)),]
mouse_hippo_pool = as.data.frame(mouse_hippo_pool)
mouse_hippo_pool$protein = row.names(mouse_hippo_pool)
mouse_hippo_pool_long = mouse_hippo_pool |>
  pivot_longer(cols = starts_with("HC2R2"),
               names_to = "sample",
               values_to = "expression") |>
  mutate(sample = factor(sample,
                         levels = paste0("HC2R2_",
                                         formatC(1:10,digits=1,flag="0")),
                         ordered = TRUE))
ggplot(data = mouse_hippo_pool_long, aes(x = sample, y = expression, color = sample)) +
  geom_boxplot() +
  xlab(NULL) +
  theme_bw() +
  scale_color_manual(values = scales::hue_pal()(length(levels(mouse_hippo_pool_long$sample)))) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = "none")
```

### Batch correcting hippocampus data
