---
title: "Data Visualization and Batch Correction"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

The previous step is [basic data wrangling](100_basic_data_wrangling.html).

### Proteome Preparation Pipeline Summary

The pipeline to summarize the proteomics data is below:

1.  Import raw peptide values.

2.  log~2~-transform and quantile normalize peptide data.

3.  Run Loess fitting on peptides within each batch against run order to correct for within-batch signal drift.

4.  Summarize peptides to protein group level (using your script).

5.  Filter out protein groups with \> 80% missing values.

6.  Run Bayesian PCA to impute missing values (needed because some batches have no non-missing samples).

7.  Use ComBat to correct for batch effects.

8.  Exclude imputed data from the batch corrected protein group dataset.

### Hippocampus Proteome Visualization

Loading in clean data.

```{r load_hippo_proteome}
adbxd_hippo_sampanno     = readRDS("./data/adbxd_hippo_sampanno_all.RDS")
adbxd_peptides_raw       = readRDS("./data/adbxd_hippo_peptides_raw.RDS")
adbxd_proteins_raw       = read.table("./data/big_report.elib.proteins.txt", sep = "\t", header = TRUE)
adbxd_peptides_annot     = readRDS("./data/adbxd_hippo_peptide_annot_raw.RDS")
```

Preparing v1.4.10 proteins

```{r proteins_prep}
adbxd_proteins_raw_annot = adbxd_proteins_raw[,1:3]
adbxd_proteins_raw_annot$UniProt = gsub("^.*\\|(.*)\\|.*_MOUSE$","\\1",adbxd_proteins_raw_annot$Protein)
row.names(adbxd_proteins_raw_annot) = adbxd_proteins_raw_annot$UniProt
adbxd_proteins_raw = as.matrix(adbxd_proteins_raw[,4:ncol(adbxd_proteins_raw)])
row.names(adbxd_proteins_raw) = adbxd_proteins_raw_annot$UniProt
adbxd_hippo_sampanno = as.data.frame(adbxd_hippo_sampanno)
row.names(adbxd_hippo_sampanno) = paste0("X", gsub("-",".",adbxd_hippo_sampanno$mzML_filename))
adbxd_proteins_raw = adbxd_proteins_raw[,row.names(adbxd_hippo_sampanno)]
colnames(adbxd_proteins_raw) = adbxd_hippo_sampanno$sample_id
```

Preparing peptides for protein group level summarization.

```{r prepare_peptides_hippo_prot_grp}
# row.names(adbxd_peptides_raw) = gsub("^(.*;.*);(\\w{2}\\|.*\\|.*)$","\\1@\\2",row.names(adbxd_peptides_raw))
```

Calling `proBatch` library and getting order of samples.

```{r}
library("proBatch")
library("ggplot2")
library("tidyverse")
adbxd_hippo_sampanno = as.data.frame(adbxd_hippo_sampanno)
row.names(adbxd_hippo_sampanno) = adbxd_hippo_sampanno$sample_id
adbxd_hippo_sampanno$FullRunName = adbxd_hippo_sampanno$sample_id
adbxd_hippo_sampanno$run_order = rank(adbxd_hippo_sampanno$creation_date)

c1_samples = adbxd_hippo_sampanno[which(adbxd_hippo_sampanno$contract == "C1"),"sample_id"]
c2_samples = adbxd_hippo_sampanno[which(adbxd_hippo_sampanno$contract == "C2"),"sample_id"]

adbxd_peptides_filtered = adbxd_peptides_raw
peptides_filter = which(rowSums(adbxd_peptides_filtered == 0) == 0)
adbxd_peptides_filtered = adbxd_peptides_filtered[peptides_filter,]
adbxd_peptides_filtered$peptide = row.names(adbxd_peptides_filtered)
```

First, visualizing the degree to which signal drift over the course of a batch is a problem.

```{r}
for (i in unique(adbxd_hippo_sampanno$batch)) {
  batch_i = adbxd_hippo_sampanno[which(adbxd_hippo_sampanno$batch == i & !is.na(adbxd_hippo_sampanno$strain)),"sample_id"]
  cors_i = cor(as.numeric(adbxd_hippo_sampanno[batch_i,"batch_elapsed"]),
               t(adbxd_peptides_filtered[,batch_i]),
               method = "spearman")
  quantile_i = quantile(cors_i, na.rm = TRUE)
  df_i = data.frame(row.names = i,
                    batch = i,
                    n = length(batch_i),
                    n_not_na_peptides = length(which(!is.na(cors_i))),
                    total_elapsed = max(as.numeric(adbxd_hippo_sampanno[batch_i,"batch_elapsed"])),
                    cors_0percentile = quantile_i["0%"],
                    cors_25percentile = quantile_i["25%"],
                    cors_50percentile = quantile_i["50%"],
                    cors_75percentile = quantile_i["75%"],
                    cors_100percentile = quantile_i["100%"])
  if (i == "C1_01") {
    batch_df = df_i
  } else {
    batch_df = rbind(batch_df, df_i)
  }
}
batch_df$contract = gsub("^(C[12])_\\d{2}$","\\1",batch_df$batch)

batch_degradation_plot = ggplot(data = batch_df, aes(x = batch, 
                            color = contract)) +
  geom_boxplot(aes(ymin = cors_0percentile,
                   lower = cors_25percentile,
                   middle = cors_50percentile,
                   upper = cors_75percentile,
                   ymax = cors_100percentile),
               stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
batch_degradation_plot
```

Visualizing raw data in hierarcical clustering

```{r adbxd_hippo_hierarc}
adbxd_hippo_log2 = log_transform_dm(as.matrix(adbxd_peptides_raw), 
                                    log_base = 2, offset = 1)
adbxd_hippo_quant = normalize_data_dm(adbxd_hippo_log2,
                                      normalize_func = "quantile")
adbxd_hippo_sampanno$batch_date = as.character(adbxd_hippo_sampanno$batch_date)
adbxd_hippo_sampanno$tissue_type = rep("Hippocampus individual", times = nrow(adbxd_hippo_sampanno))
pool_samples = grep("pool$", adbxd_hippo_sampanno$condition)
adbxd_hippo_sampanno[pool_samples,"tissue_type"] = adbxd_hippo_sampanno[pool_samples,"condition"]
adbxd_hippo_color_list = sample_annotation_to_colors(adbxd_hippo_sampanno,
                                                     factor_columns = c("batch","contract",
                                                                        "age","sex","genotype",
                                                                        "strain","batch_date",
                                                                        "tissue_type"),
                                                     numeric_columns = c("run_order","batch_elapsed"))
saveRDS(adbxd_hippo_color_list, "./data/adbxd_hippo_color_list.RDS")

plot_hierarchical_clustering(adbxd_hippo_quant,
                             sample_annotation = adbxd_hippo_sampanno,
                             color_list = adbxd_hippo_color_list,
                             factors_to_plot = c("contract","batch_date","batch",
                                                 "tissue_type","age","genotype",
                                                 "sex","strain"),
                             distance = "euclidean", agglomeration = "complete",
                             label_samples = FALSE)
```

Doing the same for the protein data

```{r visualize_protein_level}
adbxd_proteins_raw_NA_added = adbxd_proteins_raw
adbxd_proteins_raw_NA_added[which(adbxd_proteins_raw_NA_added == 0, arr.ind = TRUE)] = NA
adbxd_hippo_proteins_log2 = log_transform_dm(adbxd_proteins_raw_NA_added,
                                             log_base = 2, offset = 1)
adbxd_hippo_proteins_quant = normalize_data_dm(adbxd_hippo_proteins_log2,
                                               normalize_func = "quantile")

plot_hierarchical_clustering(adbxd_hippo_proteins_quant,
                             sample_annotation = adbxd_hippo_sampanno,
                             color_list = adbxd_hippo_color_list,
                             factors_to_plot = c("contract","batch_date","batch",
                                                 "tissue_type","age","genotype",
                                                 "sex","strain"),
                             distance = "euclidean", agglomeration = "complete",
                             label_samples = FALSE)
```
Plotting heatmap of the same.

```{r heatmap_of_quantile, fig.height = 18, fig.width = 7}
plot_heatmap_diagnostic(adbxd_hippo_quant, 
                        sample_annotation = adbxd_hippo_sampanno,
                        factors_to_plot = c("contract","batch",
                                            "tissue_type"),
                        cluster_cols = TRUE,
                        color_list = adbxd_hippo_color_list,
                        show_rownames = FALSE, show_colnames = FALSE)
```

From a discrete data perspective, the `contract` and `batch` variables are the strongest presence in the dataset.

Plotting sample mean by run order. Plotting based upon a number of variables starting with `batch`.

```{r adbxd_hippo_run_order_batch}
plot_sample_mean(adbxd_hippo_log2, adbxd_hippo_sampanno, order_col = "run_order",
                 batch_col = "batch", color_by_batch = TRUE,
                 color_scheme = adbxd_hippo_color_list[["batch"]])
```

Visualizing the effect of `contract`.

```{r adbxd_hippo_run_order_contract}
plot_sample_mean(adbxd_hippo_log2, adbxd_hippo_sampanno, order_col = "run_order",
                 batch_col = "contract", color_by_batch = TRUE,
                 color_scheme = adbxd_hippo_color_list[["contract"]])
```

Visualizing the effect of `batch_date`.

```{r adbxd_hippo_run_order_batch_date}
plot_sample_mean(adbxd_hippo_log2, adbxd_hippo_sampanno, order_col = "run_order",
                 batch_col = "batch_date", color_by_batch = TRUE,
                 color_scheme = adbxd_hippo_color_list[["batch_date"]])
```

### Batch correcting hippocampus data

The batch varibles will be dealt with next.

Starting with Loess correction within batches. Exporting this so that Loess correction may be done in parallel on the cluster.

```{r loess_correction_preparation}
# Need a column in the sample annotation sheet called "sample_id_col"
# This also needs to be a tibble
# Finally, the order column needs to be titled "order"
adbxd_hippo_sampanno$sample_id_col = adbxd_hippo_sampanno$sample_id
adbxd_hippo_sampanno$order = adbxd_hippo_sampanno$run_order

# Loess correction will be done on a total of 15 sets of columns
# These sets will be 2 or 3 batches from within the same contract
# The following grep commands create a variable to use to iterate over
# the columns in a systematic way.
adbxd_hippo_sampanno$loess_correction_n = ifelse(gsub("^C([12])_\\d{2}$","\\1",adbxd_hippo_sampanno$batch) == 1,
                                           as.numeric(gsub("^C[12]_(\\d{2})$","\\1",adbxd_hippo_sampanno$batch)),
                                           21 + as.numeric(gsub("^C[12]_(\\d{2})$","\\1",adbxd_hippo_sampanno$batch)))
adbxd_hippo_sampanno$loess_correction_n = ifelse(adbxd_hippo_sampanno$loess_correction_n < 21, (adbxd_hippo_sampanno$loess_correction_n + 1) %/% 2,
                                                 (adbxd_hippo_sampanno$loess_correction_n) %/% 2)


adbxd_hippo_sampanno_tibble = as_tibble(adbxd_hippo_sampanno)
saveRDS(adbxd_hippo_quant, "./data/adbxd_hippo_peptide_quantnorm.RDS")
saveRDS(adbxd_hippo_sampanno, "./data/adbxd_hippo_sampanno_quantnorm.RDS")
```

Note that all steps after this are performed in the next script on [batch adjustment part 2](./102_normalization_batch_adjustment_completion.html). That script has complete documentation of the Loess correction for within batch signal drift.
