---
title: "Data Batch Correction Script 2"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

The previous step is [initial batch correction](101_normalization_batch_adjustment.html).

### Hippocampus Proteome Batch Correction: Part 2

Loading in clean data.

```{r load_hippo_proteome}
library("proBatch")
library("HarmonizR")
library("ggplot2")
library("tidyverse")

adbxd_hippo_quant = readRDS("./data/adbxd_hippo_peptide_quantnorm.RDS")
adbxd_hippo_sampanno = readRDS("./data/adbxd_hippo_sampanno_quantnorm.RDS")
adbxd_hippo_sampanno_tibble = as_tibble(adbxd_hippo_sampanno)

adbxd_hippo_color_list = readRDS( "./data/adbxd_hippo_color_list.RDS")
```

The Loess smoothing step does not run in serial in any small amount of time. The script was parallelized on the cluster using the script [`./code/loess_correct_1to1500.R`](./code/loess_correct_1to1500.R). The content of this script is below (not evaluated).

**Note: the below script uses relative paths to ensure that `workflowr` does not tag it as irreproducible, but the script that runs on the cluster uses absolute paths.**

```{r loess_script_cluster, eval=FALSE}
# Loess correct: 1 to 1500
# Michael C. Saul
# michael.saul [at] jax.org
# 2022-10-05

# This script batch corrects the proteomics data in parallel.
# It splits the proteomics data matrix into 1,500 smaller chunks.
# These chunks run in parallel on the cluster.

# The script makes use of a column index and a means of producing approximately
# equal row indices.

# Getting working directory
setwd("./")

# Loading libraries
library("tidyverse")
library("tibble")
library("proBatch")

# Loading the data
adbxd_hippo_quant = readRDS("./data/adbxd_hippo_peptide_quantnorm.RDS")
adbxd_hippo_sampanno = readRDS("./data/adbxd_hippo_sampanno_quantnorm.RDS")

# Getting SLURM ID
slurm_array_id = as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))

# Getting information needed to subset the data.

# Getting rows vector
nrow_quant = nrow(adbxd_hippo_quant)
rows_slurm_id = ((slurm_array_id - 1) %/% 15) + 1
nrows_in_vector = ceiling(nrow_quant / 100)
rows_vector = (1 + (rows_slurm_id * nrows_in_vector - nrows_in_vector)):ifelse(rows_slurm_id != 100, (rows_slurm_id * nrows_in_vector), nrow_quant)

# Getting col vector
cols_slurm_id = ((slurm_array_id - 1)  %% 15) + 1

# Getting matrix slice
adbxd_hippo_sampanno_subset = adbxd_hippo_sampanno[which(adbxd_hippo_sampanno$loess_correction_n == cols_slurm_id),]
adbxd_hippo_sampanno_subset_tibble = as_tibble(adbxd_hippo_sampanno_subset)
adbxd_hippo_quant_long = matrix_to_long(adbxd_hippo_quant[rows_vector,adbxd_hippo_sampanno$sample_id])

# Running Loess regression
adbxd_hippo_loess_long = adjust_batch_trend_df(adbxd_hippo_quant_long,
                                               sample_annotation = adbxd_hippo_sampanno_subset_tibble,
                                               batch_col = "batch",
                                               order_col = "order")

adbxd_hippo_loess = long_to_matrix(adbxd_hippo_loess_long)

# Saving out data
saveRDS(adbxd_hippo_loess,paste0("./data/loess/adbxd_hippo_loess_array_",slurm_array_id,".RDS"))
```

We used the following [`./code/loess_correct_1to1500.sh`](./code/loess_correct_1to1500.sh) script to run this script:

```{sh loess_correct_shell, eval=FALSE}
#!/bin/bash
#SBATCH -p compute
#SBATCH -N 1 # number of nodes
#SBATCH -n 2 # number of cores
#SBATCH --mem=16GB # memory pool for all cores
#SBATCH -t 1-00:00 # time (D-HH:MM)
#SBATCH -o ploess_%A-%a.out # STDOUT
#SBATCH -e ploess_%A-%a.err # STDERR
#SBATCH --array=1-1500%100

# Using R v4.2.1
R_CONTAINER="/home/saulm/singularity/rocker_rstudio_4.2.1.sif"
module load singularity

singularity exec $R_CONTAINER Rscript /projects/kaczorowski-lab/USERS/saulm/adbxd_proteomics/code/loess_correct_1to1500.R
```

Now reading Loess smoothed data back in and coercing to matrix

```{r get_loess_matrix}
# Putting the matrix back together as efficiently as possible.
for (i in 1:1500) {
  adbxd_hippo_loess_i = readRDS(paste0("./data/loess/adbxd_hippo_loess_array_",i,".RDS"))
  if ((i %% 15) == 1) {
    adbxd_hippo_loess_mat_i = adbxd_hippo_loess_i
  } else {
    adbxd_hippo_loess_mat_i = cbind(adbxd_hippo_loess_mat_i, adbxd_hippo_loess_i)
  }
  if ((i %% 15) == 0) {
    if (i == 15) {
      adbxd_hippo_loess = adbxd_hippo_loess_mat_i
    } else {
      adbxd_hippo_loess = rbind(adbxd_hippo_loess, adbxd_hippo_loess_mat_i)
      rm(list = c("adbxd_hippo_loess_mat_i"))
    }
  }
}

# Reordering samples to match the quantile normalized matrix.
adbxd_hippo_loess = adbxd_hippo_loess[,adbxd_hippo_sampanno_tibble$sample_id]
adbxd_hippo_loess = adbxd_hippo_loess[row.names(adbxd_hippo_quant),]
saveRDS(adbxd_hippo_loess, "./data/adbxd_hippo_loess_peptides.RDS")

# Making a long version that can be used for tidyverse operations.
adbxd_hippo_loess_long = matrix_to_long(adbxd_hippo_loess)
adbxd_hippo_loess_long$batch = adbxd_hippo_sampanno[pull(adbxd_hippo_loess_long,FullRunName),"batch"]
adbxd_hippo_loess_long$contract = adbxd_hippo_sampanno[pull(adbxd_hippo_loess_long,FullRunName),"contract"]
```

Compressing Loess corrected data to keep things clean.

```{sh compress_loess}
# tar cvfz /projects/kacorowski-lab/USERS/saulm/adbxd_proteomics/data/loess_correction_data.tar.gz /projects/kacorowski-lab/USERS/saulm/adbxd_proteomics/data/loess
# rm -r /projects/kacorowski-lab/USERS/saulm/adbxd_proteomics/data/loess
```

Doing median centering discrete batch correction.

```{r median_correction}
# Running median correction on peptides
adbxd_hippo_median = center_feature_batch_medians_dm(adbxd_hippo_loess,
                                                     sample_annotation = adbxd_hippo_sampanno_tibble,
                                                     batch_col = "batch")
```

Evaluating how well these steps worked.

```{r evaluate_batch_correction}
plot_hierarchical_clustering(adbxd_hippo_median,
                             sample_annotation = adbxd_hippo_sampanno_tibble,
                             color_list = adbxd_hippo_color_list,
                             factors_to_plot = c("contract","batch_date","batch","sample_type",
                                                 "age","sex","genotype","strain"),
                             distance = "euclidean", agglomeration = "complete",
                             label_samples = FALSE,  fill_the_missing = NULL)
```

### Doing protein-level summarization

Using the `get_pgroup()` function as implemented by Jea Park from the MacCoss Group at UW. Using it both to run a protein group analysis with ComBat performed at the peptide level as well as ComBat performed at the protein level.

```{r get_pgroup_hippo}
library("pcaMethods")
source("./code/get_pgroup.R")

# Filtering duplicate peptides that are ambiguous
peptides = data.frame(row.names = row.names(adbxd_hippo_loess),
                      peptide_id = row.names(adbxd_hippo_loess),
                      peptide = gsub("^(.*);(.*)@(.*)$","\\1",row.names(adbxd_hippo_loess)),
                      peptide_modified = gsub("^(.*);(.*)@(.*)$","\\2",row.names(adbxd_hippo_loess)),
                      protein_id = gsub("^(.*);(.*)@(.*)$","\\3",row.names(adbxd_hippo_loess)))
duplicate_peptides = peptides |>
  group_by(peptide) |>
  summarize(n = length(unique(protein_id))) |>
  filter(n > 1) |>
  pull(peptide)
unique_peptides = row.names(peptides[which(!(peptides$peptide %in% duplicate_peptides)),])
```

Imputing peptides and doing batch correction at the peptide level

```{r impute_peptides}
adbxd_hippo_nsamples = ncol(adbxd_hippo_loess)
adbxd_hippo_loess_peptide_isna = is.na(adbxd_hippo_loess)
adbxd_hippo_peptides_keep = which(rowSums(adbxd_hippo_loess_peptide_isna) <= (0.8 * adbxd_hippo_nsamples))
adbxd_hippo_loess_filtered = adbxd_hippo_loess[adbxd_hippo_peptides_keep,]
# Running ppca, which is lower complexity than bpca
adbxd_hippo_loess_peptide_pca = pca(t(adbxd_hippo_loess_filtered),
                                    method = "ppca", nPcs = 4)
adbxd_hippo_loess_peptide_imputed = t(adbxd_hippo_loess_peptide_pca@completeObs)[row.names(adbxd_hippo_loess_filtered),colnames(adbxd_hippo_loess_filtered)]
saveRDS(adbxd_hippo_loess_peptide_imputed, "./data/adbxd_hippo_loess_peptide_imputed.RDS")

# Using ComBat to batch correct
adbxd_hippo_peptide_combat_imputed = correct_with_ComBat_dm(adbxd_hippo_loess_peptide_imputed,
                                                            sample_annotation = adbxd_hippo_sampanno_tibble,
                                                            batch_col = "batch")

# Taking imputed data out of ComBat corrected data
adbxd_hippo_peptide_combat = adbxd_hippo_peptide_combat_imputed
adbxd_hippo_peptide_combat[which(is.na(adbxd_hippo_loess_filtered), arr.ind = TRUE)] = NA

saveRDS(adbxd_hippo_peptide_combat, "./data/adbxd_hippo_peptide_combat.RDS")
```

Summarizing to protein groups and batch correcting data.

```{r summarize_to_pgroup}
# Summarizing to the protein group level
adbxd_hippo_loess_pgroup = get_pgroup(adbxd_hippo_loess[unique_peptides,],
                                      protGrp = TRUE,
                                      average = TRUE)
adbxd_hippo_loess_pgroup_isna = is.na(adbxd_hippo_loess_pgroup)

# Changing NaN to NA
adbxd_hippo_loess_pgroup[which(adbxd_hippo_loess_pgroup_isna, arr.ind = TRUE)] = NA
adbxd_hippo_nsamples = ncol(adbxd_hippo_loess_pgroup)

# Filtering protein groups with > 80% missingness
adbxd_hippo_keep = which(rowSums(adbxd_hippo_loess_pgroup_isna) <= (0.8 * adbxd_hippo_nsamples))

adbxd_hippo_loess_pgroup_filtered = adbxd_hippo_loess_pgroup[adbxd_hippo_keep,]

# Imputing missing protein groups with PCA
adbxd_hippo_loess_pgroup_pca = pca(t(adbxd_hippo_loess_pgroup_filtered), 
                                   method = "bpca", nPcs = 4)
adbxd_hippo_loess_pgroup_imputed = t(adbxd_hippo_loess_pgroup_pca@completeObs)[row.names(adbxd_hippo_loess_pgroup_filtered),colnames(adbxd_hippo_loess_pgroup_filtered)]

# Using ComBat to batch correct
adbxd_hippo_pgroup_combat_imputed = correct_with_ComBat_dm(adbxd_hippo_loess_pgroup_imputed,
                                                           sample_annotation = adbxd_hippo_sampanno_tibble,
                                                           batch_col = "batch")

# Taking imputed data out of ComBat corrected data
adbxd_hippo_pgroup_combat = adbxd_hippo_pgroup_combat_imputed
adbxd_hippo_pgroup_combat[which(is.na(adbxd_hippo_loess_pgroup_filtered), arr.ind = TRUE)] = NA
```

Looking at the protein group abundance prior to batch correction.

```{r hierarcical_clust_pgroup_prior}
plot_hierarchical_clustering(adbxd_hippo_loess_pgroup_filtered,
                             sample_annotation = adbxd_hippo_sampanno_tibble,
                             color_list = adbxd_hippo_color_list,
                             factors_to_plot = c("contract","batch_date","batch","sample_type",
                                                 "age","sex","genotype","strain"),
                             distance = "euclidean", agglomeration = "complete",
                             label_samples = FALSE, fill_the_missing = NULL,
                             plot_title = "Uncorrected Protein Groups")
```

Looking at the results of running ComBat correction after protein group summarization.

```{r hierarcical_clust_pgroup_after}
plot_hierarchical_clustering(adbxd_hippo_pgroup_combat,
                             sample_annotation = adbxd_hippo_sampanno_tibble,
                             color_list = adbxd_hippo_color_list,
                             factors_to_plot = c("contract","batch_date","batch","sample_type",
                                                 "age","sex","genotype","strain"),
                             distance = "euclidean", agglomeration = "complete",
                             label_samples = FALSE, fill_the_missing = NULL,
                             plot_title = "ComBat Corrected Protein Groups")
```

Saving protein group data.

```{r save_data}
saveRDS(adbxd_hippo_loess_pgroup_filtered,"./data/adbxd_hippo_pgroup_uncorrected.RDS")
saveRDS(adbxd_hippo_pgroup_combat_imputed,"./data/adbxd_hippo_pgroup_combat_imputed.RDS")
saveRDS(adbxd_hippo_pgroup_combat,"./data/adbxd_hippo_pgroup_combat_after.RDS")
```

The next script [visualizes data and analyzes heritability](./docs/103_heritability_analysis.html)
