---
title: "AD-BXD Resilience Analysis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

### Loading Data

Loading libraries

```{r libraries}
library("limma")
library("ggplot2")
library("tidyverse")
library("MASS")
library("qtl2")
```

Loading data.

```{r load_data}
adbxd_hippo_sampanno = readRDS("./data/adbxd_hippo_sampanno_all.RDS")
bxd_geno_data        = readRDS("./data/bxd_geno_data.RDS")
bxd_geno_meta        = readRDS("./data/bxd_geno_meta.RDS")
```

#### Calculating Behavioral Measure for Cognitive Resilience

```{r}
adbxd_5xfad_regression = adbxd_hippo_sampanno |>
  filter(condition != "Hippocampus reference pool" & condition != "Mouse brain reference pool") |>
  group_by(strain, age) |>
  mutate(cfm_arcsine = asin(sqrt(cfm_term / 100))) |>
  mutate(cfm_mean_Ntg = mean(cfm_arcsine[which(genotype == "Ntg")])) |>
  filter(genotype == "5xFAD" & !is.na(cfm_mean_Ntg)) |>
  dplyr::select(sample_id, contract, batch, strain, sex, age, genotype, cfm_term, cfm_arcsine, cfm_mean_Ntg) |>
  ungroup() |>
  group_by(strain, age) |>
  mutate(n = length(cfm_arcsine)) |>
  ungroup() |>
  as.data.frame()
```

Calculating heritability of resilience score

```{r h2_resilience_score}
library("MASS")
adbxd_5xfad_h2_df_14mo = adbxd_5xfad_regression |>
  filter(age == "Months14") |>
  mutate(obsweight = 1 / n) |>
  as.data.frame()
cfm_term_lm = lm(cfm_arcsine ~ cfm_mean_Ntg,
                 weight = obsweight,
                 data = adbxd_5xfad_h2_df_14mo)
summary(cfm_term_lm)
```

Now calculating heritability.

```{r calculate_resilience_h2}
adbxd_5xfad_h2_df_14mo$resilience = stdres(cfm_term_lm)
resilience_lm = lm(resilience ~ strain, data = adbxd_5xfad_h2_df_14mo[which(adbxd_5xfad_h2_df_14mo$n >= 2),])
resilience_h2_raw = summary(resilience_lm)$r.squared
n_per_strain = mean(adbxd_5xfad_h2_df_14mo[which(adbxd_5xfad_h2_df_14mo$n >= 2),"n"])
resilience_h2_ri = resilience_h2_raw / (resilience_h2_raw + ((1 - resilience_h2_raw) / n_per_strain))

cat("The raw heritability is ", round(resilience_h2_raw, 3), 
    " and the strain mean heritability is ", round(resilience_h2_ri, 3),
    " for the residuals-based resilience measure in 14 months aged 5xFAD mice.", sep = "")
```

Saving the resilience data

```{r}
saveRDS(adbxd_5xfad_h2_df_14mo, "./data/adbxd_5xfad_14mo_resilience.RDS")
```

Plotting the regression line.

```{r}
conf_interval_vector = data.frame(cfm_mean_Ntg = seq(from = min(adbxd_5xfad_h2_df_14mo$cfm_mean_Ntg), 
                                                     to = max(adbxd_5xfad_h2_df_14mo$cfm_mean_Ntg), 
                                                     by = 0.01))
cfm_term_lm_conf = predict(cfm_term_lm,
                           newdata = conf_interval_vector,
                           interval = "confidence", level = 0.95)
conf_interval_df = cbind(conf_interval_vector, cfm_term_lm_conf) |>
  mutate(cfm_mean_Ntg_back = 100 * (sin(cfm_mean_Ntg)) ^ 2,
         fit_back = 100 * (sin(fit)) ^ 2,
         lwr_back = 100 * (sin(lwr)) ^ 2,
         upr_back = 100 * (sin(upr)) ^ 2) |>
  as.data.frame()

adbxd_5xfad_h2_df_14mo = adbxd_5xfad_h2_df_14mo |>
  mutate(cfm_mean_Ntg_back = 100 * (sin(cfm_mean_Ntg)) ^ 2) |>
  as.data.frame()
resilience_regression_plot = ggplot(data = adbxd_5xfad_h2_df_14mo, 
                                    aes(x = cfm_mean_Ntg_back, 
                                        y = cfm_term, 
                                        size = obsweight)) + 
  geom_point(shape = 19, color = "#CC3333", alpha = 0.35) + 
  geom_line(data = conf_interval_df, 
            aes(x = cfm_mean_Ntg_back, y = fit_back),
            color = "#CC3333", size = 1) +
    geom_line(data = conf_interval_df, 
            aes(x = cfm_mean_Ntg_back, y = lwr_back),
            color = "#CC3333", size = 0.5, linetype = "dashed") +
      geom_line(data = conf_interval_df, 
            aes(x = cfm_mean_Ntg_back, y = upr_back),
            color = "#CC3333", size = 0.5, linetype = "dashed") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", size = 0.5) + 
  theme_bw() + 
  theme(panel.grid = element_line(color = "#FFFFFF"),
        legend.position = "bottom") +
  scale_size_continuous(name = "Weight", range = c(1,3)) +
  xlab("Mean Ntg 14-month CFM") +
  ylab("5xFAD 14-month CFM")
resilience_regression_plot
```

Plotting resilience score

```{r}
adbxd_5xfad_h2_df_14mo_strain_mean = adbxd_5xfad_h2_df_14mo |>
  group_by(strain) |>
  summarize(mean = mean(resilience),
            sd = sd(resilience), 
            n = length(resilience),
            cfm_mean = mean(cfm_arcsine),
            ntg_mean = mean(cfm_mean_Ntg)) |>
  mutate(sem = sd / sqrt(n),
         ul = mean + sem,
         ll = mean - sem) |>
  arrange(mean) |>
  mutate(is_bxd = gsub("^BXD.*$","BXD",strain) == "BXD",
         strain = factor(strain, levels = strain, ordered = TRUE)) |>
  mutate(straintype = ifelse(is_bxd, "BXD", as.character(strain)))

adbxd_5xfad_h2_df_14mo_strainplot = ggplot(data = adbxd_5xfad_h2_df_14mo_strain_mean, 
                                             aes(x = strain, 
                                                 y = mean,
                                                 fill = straintype)) +
  geom_col(color = "#000000") +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.5) +
  xlab("BXD Strain") +
  ylab("Resilience Score\n(Higher = More Resilient)") +
  theme_bw() +
  scale_fill_manual(values = c("#CC3333","#444444","#CCCCCC")) +
  theme(panel.grid = element_line(color = "#FFFFFF"),
        axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1),
        legend.position = "none")
adbxd_5xfad_h2_df_14mo_strainplot
```

### Mapping 5xFAD 14 month female resilience

##### Rank-Normal Function

We need a normal rank transformation function that can handle `NA` values from the following formula:

$Y^t_i={\phi}^{-1}({{r_i-c}\over{N-2c+1}})$

where $r_i$ is the rank, $N$ is the number of ranked samples, $\phi^{-1}$ is the inverse normal function (equivalent to `qnorm()` in R). A recommended value of $c=3/8$ in [Beasley and Erickson, 2010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2921808/), but we use the [van der Waerden (1952)](https://scholar.google.com/scholar_lookup?journal=Proc+Koninklijke+Nederlandse+Akademie+van+Wetenschappen&title=Order+tests+for+the+two-sample+problem+and+their+power&author=BL+van+der+Waerden&volume=55&publication_year=1952&pages=453-458&) value of $c=0$ per Vivek Philip.

Beasley and Erickson's sample size recommendation is that we need a sample of hundreds to use this function. We will attempt this with approx. 30.

```{r norm_rank_transform}
norm_rank_transform = function(x, c = 0) {
  stopifnot(is.numeric(x) & is.vector(x))
  x_noNA = which(!is.na(x))
  N = length(x_noNA)
  x[x_noNA] = qnorm((rank(x[x_noNA], ties.method = "average") - c) / (N - (2 * c) + 1))
  return(x)
}
```

##### Mapping Resilience

```{r map_5xfad_14month_female_resilience}
source("./code/gemma_bxd.R")
# Making genotype matrix

adbxd_5xfad_14mo_mapping = adbxd_5xfad_h2_df_14mo_strain_mean |>
  as.data.frame()
adbxd_genodata = cbind(data.frame(id = row.names(bxd_geno_data),
                                  x = rep("X", times = nrow(bxd_geno_data)),
                                  y = rep("Y", times = nrow(bxd_geno_data))),
                       bxd_geno_data[,as.character(adbxd_5xfad_14mo_mapping$strain)])

# Making annotation
adbxd_geno_annot = bxd_geno_meta
adbxd_geno_annot["14P_no_data","Mb_mm39"] = 3.000000
adbxd_geno_annot$id = row.names(adbxd_geno_annot)
adbxd_geno_annot$pos = adbxd_geno_annot$Mb_mm39 * 1e6
adbxd_geno_annot = adbxd_geno_annot[,c("id","pos","Chr")]
resilience_gemma = gemma_bxd(pheno = norm_rank_transform(adbxd_5xfad_14mo_mapping$mean),
                             geno = adbxd_genodata,
                             anno = adbxd_geno_annot,
                             perms = 100)
resilience_gemma_plot = plot_gemma_bxd(resilience_gemma, threshold = 0.95,
                                       genome_build = "GRCm39 Physical Coordinates") + 
  geom_hline(yintercept = resilience_gemma$gemma_perms |> 
               group_by(permu) |> 
               summarize(maxlod = max(lod)) |> 
               pull(maxlod) |> 
               quantile(probs = 1 - 0.63), 
             color = "#222222", 
             linetype = "dotted")
resilience_gemma_plot
```

Mapping the same phenotype with `R/qtl2`. Analysis inspired by the [vignette from Karl Broman from CTC 2019](https://github.com/kbroman/Teaching_CTC2019/blob/master/CTC2019_tutorial.Rmd).

```{r resilience_r_qtl2}
# Getting BXD cross data
bxd_cross_url = "https://raw.githubusercontent.com/rqtl/qtl2data/master/BXD/bxd.zip"
bxd_cross = read_cross2(bxd_cross_url)

# Calculating pseudomarkers and genoprobs
gmap = insert_pseudomarkers(bxd_cross$gmap, step=0.2, stepwidth="max")
pmap = interp_map(gmap, bxd_cross$gmap, bxd_cross$pmap)
pr = calc_genoprob(bxd_cross, gmap, error_prob=0.002, map_function="c-f")

# Calculating kinship with LOCO
k = calc_kinship(pr, "loco")

# Getting resilience phenotype
pheno = norm_rank_transform(adbxd_5xfad_14mo_mapping$mean)
names(pheno) = as.character(adbxd_5xfad_14mo_mapping$strain)

# Mapping resilience and doing permutations
out = scan1(pr, pheno, k)
n_perms = 1000
perms = scan1perm(pr, pheno, k, n_perm = n_perms)

# Plotting permutations
par(mar=c(5.1, 4.1, 0.6, 0.6))
plot_scan1(out, pmap)
abline(h = quantile(perms, 0.95))
```

We get the same result whether we use `GEMMA` or `R/qtl2`. Finding the peak location and 99% CI (in GRCm39 coordinate space).

```{r find_peak_qtl2}
library("rtracklayer")
library("liftOver")

chr8_peak = find_peaks(out, pmap, threshold = 4, prob = 0.99)
chr8_peak_granges_mm10 = GRanges(seqnames = rep(paste0("chr",chr8_peak$chr), times = 3), ranges = IRanges(start = c(chr8_peak$ci_lo, chr8_peak$pos, chr8_peak$ci_hi) * 1e6, width = c(1,1)))
names(chr8_peak_granges_mm10) = c("ci_lo","peak","ci_hi")

mm10_to_mm39_chain = import.chain("./data/raw/mm10ToMm39.over.chain")
chr8_peak_granges_mm39 = unlist(liftOver(chr8_peak_granges_mm10,
                                         chain = mm10_to_mm39_chain))
chr8_peak_mm39 = as.data.frame(chr8_peak_granges_mm39)

cat("The peak occurs at ", unique(as.character(chr8_peak_mm39$seqnames)), ":", chr8_peak_mm39["peak","start"])
cat("The 99% CI for the chr8 peak in GRCm39 coordinates spans ", unique(as.character(chr8_peak_mm39$seqnames)),":",chr8_peak_mm39["ci_lo","start"],"-",chr8_peak_mm39["ci_hi","start"],
    sep = "")
```

The chr8 peak has an empirical FDR of `r (1 + length(which(perms >= chr8_peak[1,"lod"]))) / n_perms`.

Checking to see if we would've gotten the same result by mapping the `asin(sqrt())` transforms of 5xFAD and Ntg CFM.

Starting with Ntg.

```{r plot_ntg_cfm_mapping}
adbxd_ntg_14mo_cfm = adbxd_hippo_sampanno |>
  filter(condition != "Hippocampus reference pool" & 
         condition != "Mouse brain reference pool") |>
  mutate(cfm_arcsine = asin(sqrt(cfm_term / 100))) |>
  dplyr::select(sample_id, contract,  batch, 
                strain, sex, age, genotype, 
                cfm_term, cfm_arcsine) |>
  filter(age == "Months14" & 
         genotype == "Ntg" & 
         !(strain %in% c("C57BL/6J","DBA/2J"))) |>
  group_by(strain) |>
  summarize(mean = mean(cfm_arcsine, na.rm = TRUE),
            sd = sd(cfm_arcsine, na.rm = TRUE),
            n = length(which(!is.na(cfm_arcsine)))) |>
  mutate(sem = sd / sqrt(n), 
         ll = mean - sem,
         ul = mean + sem)
adbxd_ntg_genodata = cbind(data.frame(id = row.names(bxd_geno_data),
                                      x = rep("X", times = nrow(bxd_geno_data)),
                                      y = rep("Y", times = nrow(bxd_geno_data))),
                           bxd_geno_data[,as.character(adbxd_ntg_14mo_cfm$strain)])

cfm_ntg_gemma = gemma_bxd(pheno = norm_rank_transform(adbxd_ntg_14mo_cfm$mean),
                             geno = adbxd_ntg_genodata,
                             anno = adbxd_geno_annot,
                             perms = 100)
cfm_ntg_gemma_plot = plot_gemma_bxd(cfm_ntg_gemma, threshold = 0.95,
                                      colors = c("#444455","#444455"),
                                      genome_build = "GRCm39 Physical Coordinates") + 
  geom_hline(yintercept = cfm_ntg_gemma$gemma_perms |> 
               group_by(permu) |> 
               summarize(maxlod = max(lod)) |> 
               pull(maxlod) |> 
               quantile(probs = 1 - 0.63), 
             color = "#222222", 
             linetype = "dotted")
cfm_ntg_gemma_plot
```

Doing the same for 5xFAD.

```{r plot_5xfad_cfm_mapping}
adbxd_5xfad_14mo_cfm = adbxd_hippo_sampanno |>
  filter(condition != "Hippocampus reference pool" & 
         condition != "Mouse brain reference pool") |>
  mutate(cfm_arcsine = asin(sqrt(cfm_term / 100))) |>
  dplyr::select(sample_id, contract,  batch, 
                strain, sex, age, genotype, 
                cfm_term, cfm_arcsine) |>
  filter(age == "Months14" & 
         genotype == "5xFAD" & 
         !(strain %in% c("C57BL/6J","DBA/2J"))) |>
  group_by(strain) |>
  summarize(mean = mean(cfm_arcsine, na.rm = TRUE),
            sd = sd(cfm_arcsine, na.rm = TRUE),
            n = length(which(!is.na(cfm_arcsine)))) |>
  mutate(sem = sd / sqrt(n), 
         ll = mean - sem,
         ul = mean + sem)
adbxd_5xfad_genodata = cbind(data.frame(id = row.names(bxd_geno_data),
                                      x = rep("X", times = nrow(bxd_geno_data)),
                                      y = rep("Y", times = nrow(bxd_geno_data))),
                           bxd_geno_data[,as.character(adbxd_5xfad_14mo_cfm$strain)])

cfm_5xfad_gemma = gemma_bxd(pheno = norm_rank_transform(adbxd_5xfad_14mo_cfm$mean),
                             geno = adbxd_5xfad_genodata,
                             anno = adbxd_geno_annot,
                             perms = 100)
cfm_5xfad_gemma_plot = plot_gemma_bxd(cfm_5xfad_gemma, threshold = 0.95, 
                                      colors = c("#CC3333","#CC3333"),
                                      genome_build = "GRCm39 Physical Coordinates") + 
  geom_hline(yintercept = cfm_5xfad_gemma$gemma_perms |> 
               group_by(permu) |> 
               summarize(maxlod = max(lod)) |> 
               pull(maxlod) |> 
               quantile(probs = 1 - 0.63), 
             color = "#222222", 
             linetype = "dotted")
cfm_5xfad_gemma_plot
```

Doing 5xFAD genotype by BXD genotype interaction mapping

```{r plot_interact_cfm_mapping}
adbxd_14mo_cfm = adbxd_hippo_sampanno |>
  filter(condition != "Hippocampus reference pool" & 
         condition != "Mouse brain reference pool") |>
  mutate(cfm_arcsine = asin(sqrt(cfm_term / 100))) |>
  dplyr::select(sample_id, contract,  batch, 
                strain, sex, age, genotype, 
                cfm_term, cfm_arcsine) |>
  filter(age == "Months14" & 
         !(strain %in% c("C57BL/6J","DBA/2J"))) |>
  group_by(strain, genotype) |>
  summarize(mean = mean(cfm_arcsine, na.rm = TRUE),
            sd = sd(cfm_arcsine, na.rm = TRUE),
            n = length(which(!is.na(cfm_arcsine)))) |>
  mutate(sem = sd / sqrt(n), 
         ll = mean - sem,
         ul = mean + sem) |>
  as.data.frame()

adbxd_inter_genodata = cbind(data.frame(id = row.names(bxd_geno_data),
                                      x = rep("X", times = nrow(bxd_geno_data)),
                                      y = rep("Y", times = nrow(bxd_geno_data))),
                           bxd_geno_data[,as.character(adbxd_14mo_cfm$strain)])

cfm_inter_gemma = gemma_bxd(pheno = adbxd_14mo_cfm$mean,
                             geno = adbxd_inter_genodata,
                             anno = adbxd_geno_annot,
                            gxe = ifelse(adbxd_14mo_cfm$genotype == "5xFAD", 1, 0),
                            loco = FALSE,
                             perms = 100)
cfm_inter_gemma_plot = plot_gemma_bxd(cfm_inter_gemma, threshold = 0.95) +
  geom_hline(yintercept = cfm_inter_gemma$gemma_perms |> 
               group_by(permu) |> 
               summarize(maxlod = max(lod)) |> 
               pull(maxlod) |> 
               quantile(probs = 1 - 0.63), 
             color = "#222222", 
             linetype = "dotted")
cfm_inter_gemma_plot
```

Showing the top marker on chr8.

```{r}
library("ggbeeswarm")
adbxd_5xfad_14mo_mapping$rs33539160 = ifelse(as.character(adbxd_genodata["rs33539160",as.character(adbxd_5xfad_14mo_mapping$strain)]) == 0, "B6/B6", "B6/D2")
rs33539160_plot = ggplot(data = adbxd_5xfad_14mo_mapping, 
       aes(x = rs33539160, 
           y = mean, 
           color = rs33539160)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_beeswarm() +
  theme_bw() +
  theme(panel.grid = element_line(color = "#FFFFFF"),
        legend.position = "bottom") +
  scale_color_manual(values = c("#000000","#999999")) +
  ylab("Strain Mean Resilience Score")
rs33539160_plot
```

Making a figure of the results.

```{r figure_1}
library("cowplot")
source("./code/ggplot_pub_themes.R")
top_row = plot_grid(resilience_regression_plot + 
                      pubtheme_bw +
                      theme(legend.position = "bottom"), 
                    adbxd_5xfad_h2_df_14mo_strainplot + 
                      pubtheme_bw + 
                      theme(legend.position = "none",
                            axis.text.x = element_text(angle = 45,
                                                       vjust = 1,
                                                       hjust = 1)), 
                    labels = c("A","B"), 
                    label_size = 12, ncol = 2, 
                    rel_widths = c(1,1.75),
                    align = "h", axis = "b")
# top_row
bottom_rows = plot_grid(resilience_gemma_plot + 
                          xlab(NULL) + 
                          pubtheme_bw + 
                          theme(legend.position = "none"),
                        cfm_ntg_gemma_plot + 
                          xlab(NULL) + 
                          pubtheme_bw + 
                          theme(legend.position = "none"),
                        cfm_5xfad_gemma_plot + 
                          pubtheme_bw + 
                          theme(legend.position = "none"),
                        labels = c("C","D","E"),
                        label_size = 12, nrow = 3,
                        rel_heights = c(1.5,1,1),
                        align = "v", axis = "b")
# bottom_rows
figure_1 = plot_grid(top_row, bottom_rows, nrow = 2, rel_heights = c(1,2))
figure_1
saveRDS(figure_1, "./data/figure_1.RDS")
ggsave("./output/figure_1.pdf", figure_1, width = 174, height = 174, units = "mm")
```
