---
title: "AD-BXD Proteome Abundance and Mediation Analysis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

### Loading Data

Loading libraries

```{r libraries}
library("limma")
library("qvalue")
library("ggplot2")
library("ggbeeswarm")
library("ggrepel")
library("ggtext")
library("cowplot")
library("cliqueR")
library("tidyverse")
library("MASS")
library("qtl2")
library("pcaMethods")
library("topGO")
source("./code/ggplot_pub_themes.R")
```

Loading proteome and genotype data.

```{r load_proteome_data}
adbxd_hippo_pgroup            = readRDS("./data/adbxd_hippo_pgroup_annotated.RDS")
adbxd_hippo_sampanno          = readRDS("./data/adbxd_hippo_sampanno_quantnorm.RDS")
adbxd_hippo_pgroup_annotation = readRDS("./data/adbxd_hippo_pgroup_annotation.RDS")
adbxd_annot_list_coordinates  = readRDS("./data/adbxd_uniprot_coordinates.RDS")
adbxd_annot_list_uniprot      = readRDS("./data/adbxd_uniprot_alldata.RDS")
adbxd_coord_df                = readRDS("./data/adbxd_uniprot_coord_df.RDS")
adbxd_genotypes               = readRDS("./data/bxd_geno_data.RDS")
adbxd_geno_meta               = readRDS("./data/bxd_geno_meta.RDS")
adbxd_5xfad_14mo_resilience   = readRDS("./data/adbxd_5xfad_14mo_resilience.RDS")
row.names(adbxd_5xfad_14mo_resilience) = adbxd_5xfad_14mo_resilience$sample_id
```

Getting rs33539160 genotype in the sample annotation data and filtering for 14-month-old 5xFAD animals.

```{r}
adbxd_hippo_sampanno$rs33539160 = ifelse(!is.na(adbxd_hippo_sampanno$strain),
                                         as.data.frame(t(adbxd_genotypes))[adbxd_hippo_sampanno$strain,"rs33539160"],
                                         NA)
adbxd_hippo_sampanno_5xfad_months14 = adbxd_hippo_sampanno |>
  filter(age == "Months14" & genotype == "5xFAD") |>
  mutate(rs33539160 = ifelse(rs33539160 == 1, "B6/D2","B6/B6")) |>
  as.data.frame()

adbxd_hippo_pgroup_5xfad_months14 = adbxd_hippo_pgroup[,row.names(adbxd_hippo_sampanno_5xfad_months14)]
adbxd_hippo_pgroup_5xfad_months14 = adbxd_hippo_pgroup_5xfad_months14[which(rowSums(is.na(adbxd_hippo_pgroup_5xfad_months14)) < 0.2 * ncol(adbxd_hippo_pgroup_5xfad_months14)),]
```

Plotting genotype effect for resilience at rs33539160.

```{r}
library("ggbeeswarm")
adbxd_5xfad_14mo_resilience$rs33539160 = ifelse(!is.na(adbxd_5xfad_14mo_resilience$strain),
                                         as.data.frame(t(adbxd_genotypes))[adbxd_5xfad_14mo_resilience$strain,"rs33539160"],
                                         NA)
rs33539160_plot = adbxd_5xfad_14mo_resilience |>
  filter(!is.na(rs33539160)) |>
  mutate(strain = as.character(strain),
         rs33539160 = ifelse(rs33539160 == 1, "B6/D2", "B6/B6")) |>
  group_by(strain, rs33539160) |>
  summarize(mean = mean(resilience, na.rm = TRUE)) |>
  ggplot(aes(x = rs33539160, 
             y = mean, 
             color = rs33539160)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_beeswarm(size = 0.25) +
  # ggtitle("") +
  theme_bw() +
  pubtheme_bw + 
  theme(panel.grid = element_line(color = "#FFFFFF"),
        legend.position = "none") +
  scale_color_manual(values = c("#000000","#999999")) +
  ylab("Strain Mean Resilience Score")
rs33539160_plot
```

Running differential abundance analysis on 14-month 5xFAD animals.

```{r}
# Getting model matrix
adbxd_modmat = model.matrix(~ sex + rs33539160,
                            data = adbxd_hippo_sampanno_5xfad_months14)

# Running model
adbxd_month14_5xfad_rs33539160_lmFit = lmFit(adbxd_hippo_pgroup_5xfad_months14, 
                                             design = adbxd_modmat, 
                                             method = "robust", weights = NULL)
adbxd_month14_5xfad_rs33539160_lmFit = eBayes(adbxd_month14_5xfad_rs33539160_lmFit, trend = TRUE, robust = TRUE)

# Compiling top tables
adbxd_rs33539160_toptable = as.data.frame(topTable(adbxd_month14_5xfad_rs33539160_lmFit,
                                               coef = grep("^rs33539160", colnames(adbxd_modmat)), 
                                               number = nrow(adbxd_hippo_pgroup_5xfad_months14)))
adbxd_rs33539160_toptable$qvalue = qvalue(adbxd_rs33539160_toptable$P.Value)$qvalues
saveRDS(adbxd_rs33539160_toptable, "./data/adbxd_rs33539160_toptable.RDS")
```

Showing top table for the proteins at q < 0.10.

```{r proteins_q_0p10}
adbxd_rs33539160_toptable |>
  filter(qvalue < 0.10)
```

Plotting a volcano plot of the genotype data.

```{r}
y_threshold = mean(c(-1 * log10(max(adbxd_rs33539160_toptable[which(adbxd_rs33539160_toptable$qvalue < 0.01),"P.Value"])),
                     -1 * log10(min(adbxd_rs33539160_toptable[which(adbxd_rs33539160_toptable$qvalue > 0.01),"P.Value"]))))

adbxd_rs33539160_toptable$protein = row.names(adbxd_rs33539160_toptable
                                              )
adbxd_rs33539160_volcano_plot = adbxd_rs33539160_toptable |>
  mutate(log10pvalue = -1 * log10(P.Value),
         label = ifelse(protein == "P47791", "GSHR", ""),
         label = ifelse(protein == "P46096", "SYT1", label),
         label = ifelse(protein == "O54829", "RGS14", label),
         color = ifelse(label == "", "not", "sig"),
         color = ifelse(protein == "P47791", "GSHR", color)) |>
  ggplot(aes(x = logFC, y = log10pvalue, label = label, color = color)) +
  geom_hline(yintercept = y_threshold, linetype = "dashed", color = "#555555") +
  # geom_vline(xintercept = -0.5, linetype = "dashed", color = "#555555") +
  # geom_vline(xintercept = 0.5, linetype = "dashed", color = "#555555") +
  geom_point(size = 0.25) +
  geom_label_repel(size = 6 * pt_to_mm) +
  scale_color_manual(values = c("#CC3333","#AAAAAA","#000000")) + 
  xlab("log<sub>2</sub>(Fold-Change) for rs33539160<br />Higher = More Expression in B6/D2 Genotype") +
  ylab("-log<sub>10</sub>(p-value)") +
  # ggtitle("") +
  theme_bw() +
  pubtheme_bw + 
  theme(legend.position = "none",
        panel.grid = element_line(color = "#FFFFFF"),
        axis.title.x = element_markdown(),
        axis.title.y = element_markdown())

adbxd_rs33539160_volcano_plot
```

Plotting GSHR

```{r plot_GSHR}
adbxd_5xfad_14mo_resilience$GSHR = adbxd_hippo_pgroup_5xfad_months14["P47791",row.names(adbxd_5xfad_14mo_resilience)]

GSHR_plot = adbxd_5xfad_14mo_resilience |>
  filter(!is.na(rs33539160)) |>
  mutate(strain = as.character(strain),
         rs33539160 = ifelse(rs33539160 == 1, "B6/D2", "B6/B6")) |>
  group_by(strain, rs33539160) |>
  summarize(mean = mean(GSHR, na.rm = TRUE)) |>
  ggplot(aes(x = rs33539160, 
             y = mean, 
             color = rs33539160)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_beeswarm(size = 0.25) +
  # ggtitle("") +
  theme_bw() +
  pubtheme_bw +
  theme(panel.grid = element_line(color = "#FFFFFF"),
        legend.position = "none") +
  scale_color_manual(values = c("#000000","#999999")) +
  ylab("Strain Mean GSHR Expression")
GSHR_plot
```

Building the expression figure.

```{r}
figure_2 = plot_grid(rs33539160_plot, 
                     adbxd_rs33539160_volcano_plot,
                     GSHR_plot,
                     labels = LETTERS[1:3],
                     ncol = 3, 
                     rel_widths = c(1,2,1),
                     label_size = 12,
                     align = "h")
figure_2
saveRDS(figure_2, "./data/figure_2.RDS")
ggsave("./output/figure_2.pdf", figure_2, width = 174, height = 174 / 3, units = "mm")
```

Doing regression on resilience.

```{r resilience_regression}
adbxd_5xfad_14mo_resilience = adbxd_5xfad_14mo_resilience[colnames(adbxd_hippo_pgroup_5xfad_months14),]
adbxd_resilience_modmat = model.matrix(~ sex + resilience,
                                       data = adbxd_5xfad_14mo_resilience)

# Running model
adbxd_resilience_lmFit = lmFit(adbxd_hippo_pgroup_5xfad_months14, 
                                             design = adbxd_resilience_modmat, 
                                             method = "robust", weights = NULL)
adbxd_resilience_lmFit = eBayes(adbxd_resilience_lmFit, trend = TRUE, robust = TRUE)

# Compiling top tables
adbxd_resilience_toptable = as.data.frame(topTable(adbxd_resilience_lmFit,
                                               coef = grep("^resilience", colnames(adbxd_resilience_modmat)), 
                                               number = nrow(adbxd_hippo_pgroup_5xfad_months14)))
adbxd_resilience_toptable$qvalue = qvalue(adbxd_resilience_toptable$P.Value)$qvalues
saveRDS(adbxd_resilience_toptable, "./data/adbxd_resilience_months14_toptable.RDS")
```

Displaying top resilience-correlated proteins.

```{r resilience_proteins_q_0p10}
adbxd_resilience_toptable |>
  filter(qvalue < 0.10)
```
